---
title: "A quick start guide to rTrafa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A quick start guide to rTrafa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette provides a quick start guide to get up and running with rTrafa as fast as possible. For a more comprehensive introduction, see [Introduction to rTrafa](introduction-to-rtrafa.html).

```{r install, eval = FALSE}
install.packages("rTrafa")
```

```{r setup}
library("rTrafa")
```

In this guide we walk you through five steps to discover products, explore their structure, download data and visualise it.

Trafa's data is organised in three layers: **Product** (a statistical domain, e.g. "Buses"), **Measure** (what is measured, e.g. "Vehicles in traffic"), and **Dimension** (how to filter, e.g. year, fuel type). To download data you specify a product, a measure, and optionally dimension filters.

### 1. Get products

Trafa publishes dozens of statistical products. Start by listing them:

```{r, echo = FALSE}
products <- rTrafa:::vd_products
```

```{r, eval = FALSE}
products <- get_products()
```

```{r}
dplyr::glimpse(products)
```

Use `product_search()` to filter by text:

```{r}
products |>
  product_search("Bussar") |>
  product_describe(max_n = 3)
```

### 2. Explore measures

Each product has several measures (KPIs). Let's see what's available for Buses (t10011):

```{r, echo = FALSE}
measures <- rTrafa:::vd_measures_bussar
```

```{r, eval = FALSE}
measures <- get_measures("t10011")
```

```{r}
measures |> measure_describe()
```

### 3. Explore dimensions

Dimensions are the variables you can filter on. Use `get_dimensions()` to discover them:

```{r, echo = FALSE}
dims <- rTrafa:::vd_dims_bussar
```

```{r, eval = FALSE}
dims <- get_dimensions("t10011")
```

```{r}
dims |> dimension_describe()
```

Notice that some dimensions belong to a **hierarchy** (shown in brackets). These are grouped conceptually but each is queried independently.

You can inspect the available values for any dimension:

```{r, echo = FALSE}
ar_vals <- rTrafa:::vd_ar_values
```

```{r, eval = FALSE}
ar_vals <- dimension_values(dims, "ar")
```

```{r}
ar_vals
```

Note the `type` column: `"filter"` entries like `senaste` (latest) and `forra` (previous) are shortcuts that the API resolves dynamically.

### 4. Get data

Now let's fetch data. We request the number of buses in traffic for the last 10 years:

```{r, echo = FALSE}
bus_data <- rTrafa:::vd_data_bussar
```

```{r, eval = FALSE}
bus_data <- get_data("t10011", "itrfslut",
  ar = as.character(2016:2025)
)
```

```{r}
dplyr::glimpse(bus_data)
```

### 5. Visualise

Note how we convert the `ar` column to a proper `Date` before plotting.
Trafa returns `ar` as a character column (`"2016"`, `"2017"`, ...), and
plotting it as an integer can produce awkward `ggplot2` breaks like
`2020, 2022.5, 2025`. Wrapping it in `as.Date(paste0(ar, "-01-01"))`
lets `scale_x_date()` place tick marks on whole years — a pattern
worth reusing for any time-series analysis.

```{r, fig.width = 6, fig.height = 4}
library("ggplot2")

bus_plot <- bus_data |>
  dplyr::mutate(year = as.Date(paste0(ar, "-01-01")))

ggplot(bus_plot, aes(x = year, y = itrfslut)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Buses in traffic in Sweden",
    x = "Year",
    y = "Number of vehicles",
    caption = data_legend(bus_data)
  ) +
  theme_minimal()
```

## Next steps

- **Deeper walkthrough** — `vignette("introduction-to-rtrafa")` covers
  the full data model, hierarchies, filter shortcuts, and dimension
  validation.
- **Filter shortcuts** — Use `ar = "senaste"` to always get the latest
  year without hardcoding.

## Related packages

`rTrafa` is part of a family of R packages for Swedish and Nordic
open statistics that share the same design philosophy:

- [rKolada](https://lchansson.github.io/rKolada/) — R client for the
  [Kolada](https://kolada.se/) database of Swedish municipal and
  regional Key Performance Indicators
- [pixieweb](https://lchansson.github.io/pixieweb/) — R client for
  PX-Web APIs (Statistics Sweden, Statistics Norway, Statistics Finland,
  and more)
