---
title: "Introduction to rTrafa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to rTrafa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`rTrafa` is an R package for *discovering*, *inspecting* and *downloading* Swedish transport statistics from the [Trafa API](https://api.trafa.se/). This vignette provides an overview of the methods included in the `rTrafa` package and the design principles of the package API. To learn more about the specifics of functions and to see a full list of the functions included, please see the [Reference section of the package homepage](https://lchansson.github.io/rTrafa/reference/index.html) or run `??rTrafa`. For a quick introduction to the package see the vignette [A quick start guide to rTrafa](a-quickstart-rtrafa.html).

The design of `rTrafa` functions is inspired by the design and functionality provided by several packages in the `tidyverse` family. rTrafa uses the base R pipe (`|>`) throughout. Some vignette examples use `dplyr` and `ggplot2` for data wrangling and visualisation:

```{r install, eval = FALSE}
install.packages("rTrafa")
```

```{r setup}
library("rTrafa")
```

## Trafa's data model

The Trafa API organises Swedish transport statistics in a four-level hierarchy:

1. **Product** — A statistical domain, e.g. *Buses* (t10011) or *Passenger cars* (t10016). Think of it as a family of related statistics.
2. **Measure** — A specific KPI within a product. Buses has four measures: vehicles in traffic, deregistered, newly registered, and deregistrations.
3. **Dimension** — A filterable variable such as year, fuel type, or owner category. Each dimension has discrete values you can select.
4. **Value** — The individual codes within a dimension (e.g. `"2024"`, `"102"` for Diesel).

There is also a grouping concept called **Hierarchy**. A hierarchy like "Owner" groups related dimensions ("Owner category" and "Licence type") together. Hierarchies are purely organisational — they cannot be used directly in queries. Their child dimensions appear in `get_dimensions()` with the hierarchy name noted in the `hierarchy` column.

rTrafa provides a function family for each of the first three levels:

| Level | List | Search | Describe | Extract | Values |
|-------|------|--------|----------|---------|--------|
| Product | `get_products()` | `product_search()` | `product_describe()` | `product_extract_ids()` | — |
| Measure | `get_measures()` | `measure_search()` | `measure_describe()` | `measure_extract_names()` | — |
| Dimension | `get_dimensions()` | `dimension_search()` | `dimension_describe()` | `dimension_extract_names()` | `dimension_values()` |

## Products

```{r, echo = FALSE}
products <- rTrafa:::vd_products
```

```{r, eval = FALSE}
products <- get_products()
```

The product table has one row per statistical product:

```{r}
dplyr::glimpse(products)
```

All entity tables support the same set of operations. `product_search()` filters by regex:

```{r}
products |>
  product_search("Personbilar") |>
  product_describe(max_n = 2)
```

`product_extract_ids()` extracts the `name` column as a character vector — useful for passing into downstream functions:

```{r}
products |>
  product_search("Personbilar") |>
  product_extract_ids()
```

## Measures

Each product contains one or more measures. A measure is the statistic being observed — for example, *Antal i trafik* (vehicles in traffic) or *Antal nyregistreringar* (new registrations).

```{r, echo = FALSE}
measures <- rTrafa:::vd_measures_bussar
```

```{r, eval = FALSE}
measures <- get_measures("t10011")
```

```{r}
measures |> measure_describe()
```

Extract the names to use in `get_data()`:

```{r}
measures |> measure_extract_names()
```

## Dimensions

Dimensions are the filter variables available for a product. Use `get_dimensions()` to discover them:

```{r, echo = FALSE}
dims <- rTrafa:::vd_dims_bussar
```

```{r, eval = FALSE}
dims <- get_dimensions("t10011")
```

```{r}
dims |> dimension_describe()
```

### Hierarchies

Some dimensions are grouped under a hierarchy. In the output above, dimensions marked with `[agare]` belong to the "Owner" hierarchy. This is metadata that helps you understand the logical structure — but when building a query, you use the child dimension names directly (e.g. `agarkat`, not `agare`).

```{r}
dims |>
  dplyr::filter(!is.na(hierarchy))
```

### Dimension values and filter shortcuts

Each dimension has a set of valid values. `dimension_values()` returns them as a tibble:

```{r, echo = FALSE}
ar_vals <- rTrafa:::vd_ar_values
```

```{r, eval = FALSE}
ar_vals <- dimension_values(dims, "ar")
```

```{r}
ar_vals
```

Note the `type` column. Entries with `type = "filter"` are **filter shortcuts** — dynamic values that the API resolves at query time:

- `senaste` — the most recent available value
- `forra` — the previous value

These are useful for building queries that stay current without hardcoding years:

```{r, eval = FALSE}
# Always gets the latest year, regardless of when you run it
get_data("t10011", "itrfslut", ar = "senaste")
```

Regular dimension values (`type = "value"`) can be inspected too:

```{r, echo = FALSE}
drivm_vals <- rTrafa:::vd_drivm_values
```

```{r, eval = FALSE}
drivm_vals <- dimension_values(dims, "drivm")
```

```{r}
drivm_vals
```

### Dimension validation

When you know which measure you want, `get_dimensions()` can validate which dimensions are compatible:

```{r, eval = FALSE}
# Only dimensions valid for the "itrfslut" measure
get_dimensions("t10011", measure = "itrfslut")

# See all dimensions with their validity status
get_dimensions("t10011", measure = "itrfslut", only_valid = FALSE)
```

You can also pass **several measures at once**. The API returns the intersection — only dimensions that are valid for *every* measure in the vector. This is useful when you want to compare or combine measures and need to know which filters you can apply uniformly:

```{r, eval = FALSE}
# Dimensions valid for BOTH "vehicles in traffic" and "new registrations"
get_dimensions("t10011", measure = c("itrfslut", "nyregunder"))
```

For Buses, only `ar` (year) survives this intersection — most other dimensions are specific to one measure or the other.

## Fetching data

`get_data()` is the core function. Provide a product, a measure, and optional dimension filters:

```{r, echo = FALSE}
bus_data <- rTrafa:::vd_data_bussar
```

```{r, eval = FALSE}
bus_data <- get_data("t10011", "itrfslut",
  ar = as.character(2016:2025)
)
```

```{r}
dplyr::glimpse(bus_data)
```

When `simplify = TRUE` (the default), each dimension gets both a code column and a `_label` column with the human-readable name.

### Prepared queries

For complex queries, `prepare_query()` validates your selections before hitting the data endpoint:

```{r, eval = FALSE}
q <- prepare_query("t10011", "itrfslut",
  ar = c("2023", "2024"),
  drivm = c("101", "102", "103")
)
q
#> Trafa query: t10011 | itrfslut
#>   Dimension filters:
#>     ar = c("2023", "2024")
#>     drivm = c("101", "102", "103")
#>   Available (unfiltered) dimensions: ...

# Pass the validated query to get_data()
get_data(query = q)
```

### Data helpers

`data_minimize()` removes columns where all values are identical — useful after filtering to a single year or fuel type:

```{r, eval = FALSE}
get_data("t10011", "itrfslut", ar = "2024") |>
  data_minimize()
```

`data_legend()` generates a source caption for plots:

```{r}
data_legend(bus_data)
```

## Example: buses in traffic over time

Note how we convert the `ar` column to a proper `Date` before plotting.
Trafa returns `ar` as a character column (`"2016"`, `"2017"`, ...), and
plotting it as an integer can produce awkward `ggplot2` breaks like
`2020, 2022.5, 2025`. Wrapping it in `as.Date(paste0(ar, "-01-01"))`
lets `scale_x_date()` place tick marks on whole years. This is a pattern
you'll want to reuse for any time-series analysis — both in `rTrafa`
and in the sibling packages `rKolada` and `pixieweb`.

```{r, fig.width = 6, fig.height = 4}
library("ggplot2")

bus_plot <- bus_data |>
  dplyr::mutate(year = as.Date(paste0(ar, "-01-01")))

ggplot(bus_plot, aes(x = year, y = itrfslut)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  # Years as dates, one tick per year
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Buses in traffic in Sweden",
    x = "Year",
    y = "Number of vehicles",
    caption = data_legend(bus_data)
  ) +
  theme_minimal()
```

## Example: fuel type breakdown

```{r, echo = FALSE}
drivm_data <- rTrafa:::vd_data_drivm
```

```{r, eval = FALSE}
drivm_data <- get_data("t10011", "itrfslut",
  ar = "senaste",
  drivm = c("101", "102", "103", "104", "105", "106", "107")
)
```

```{r, fig.width = 7, fig.height = 4}
ggplot(drivm_data, aes(
  x = reorder(drivm_label, -itrfslut),
  y = itrfslut
)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Buses in traffic by fuel type",
    subtitle = paste("Year:", drivm_data$ar_label[1]),
    x = NULL,
    y = "Number of vehicles",
    caption = data_legend(drivm_data)
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

## Example: comparing vehicle types

By fetching the same measure from different products we can compare across vehicle types:

```{r, echo = FALSE}
pbilar_data <- rTrafa:::vd_data_personbilar
```

```{r, eval = FALSE}
pbilar_data <- get_data("t10016", "itrfslut",
  ar = as.character(2016:2025)
)
```

```{r, fig.width = 7, fig.height = 4}
comparison <- dplyr::bind_rows(
  bus_data |> dplyr::mutate(vehicle = "Buses"),
  pbilar_data |> dplyr::mutate(vehicle = "Passenger cars")
) |>
  dplyr::mutate(year = as.Date(paste0(ar, "-01-01")))

ggplot(comparison, aes(x = year, y = itrfslut, color = vehicle)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Vehicles in traffic in Sweden",
    x = "Year",
    y = "Number of vehicles",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

## Related packages

`rTrafa` is part of a family of R packages for Swedish and Nordic
open statistics that share the same design philosophy — tibble-based,
pipe-friendly, and offline-safe:

- [rKolada](https://lchansson.github.io/rKolada/) — R client for the
  [Kolada](https://kolada.se/) database of Swedish municipal and
  regional Key Performance Indicators
- [pixieweb](https://lchansson.github.io/pixieweb/) — R client for
  PX-Web APIs (Statistics Sweden, Statistics Norway, Statistics Finland,
  and more)
